A Linear Inside-Outside Algorithm for Correcting Sequencing Errors in Structured RNA

### 2013-01-09 Recomb/main_RECOMB.tex
format llncs

\documentclass[11pt]{article}
\usepackage[top=1in, bottom=1in,right=1in,left=1in]{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm} %pour les newtheorem
\usepackage{epstopdf}
\usepackage{xspace}
\usepackage{relsize}
\usepackage[numbers]{natbib}
\usepackage[colorinlistoftodos]{todonotes}

%\usepackage{subfig} %Subfigures
\usepackage{subcaption} %Subfigures

@@

%\title{An inside-outside algorithms for mutant RNAs with applications to error detections in structured RNAs}
\title{A linear inside-outside algorithm for correcting sequencing errors in structured RNA}
\author{Vladimir Reinharz, Yann Ponty$^{*}$, J\'er\^{o}me Waldisp\"{u}hl$^*$}
\date{}

>>>>

\documentclass{llncs}
%\usepackage[top=1in, bottom=1in,right=1in,left=1in]{geometry}                % See geometry.pdf to learn the layout options. There are lots.
%\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
%\usepackage{amsthm} %pour les newtheorem
\usepackage{epstopdf}
\usepackage{xspace}
\usepackage{relsize}
\usepackage[numbers]{natbib}
\usepackage[colorinlistoftodos]{todonotes}

%\usepackage{subfig} %Subfigures
%\usepackage{subcaption} %Subfigures

@@

\newcommand{\Ab}{{\sf{A}}}
\newcommand{\Cb}{{\sf{C}}}
\newcommand{\Gb}{{\sf{G}}}
\newcommand{\Ub}{{\sf{U}}}

%\title{An inside-outside algorithms for mutant RNAs with applications to error detections in structured RNAs}
\title{A Linear Inside-Outside Algorithm for Correcting Sequencing Errors in Structured RNA}
\author{Vladimir Reinharz\inst{1}, Yann Ponty\inst{2}$^{*}$ \and J\'er\^{o}me Waldisp\"{u}hl\inst{1}$^*$}
\date{}
\institute{School of Computer Science, McGill University, Montreal, Canada.
	\and  Laboratoire d'informatique, \'Ecole Polytechnique, Palaiseau, France.
	 \\\email{jeromew@cs.mcgill.ca}, \email{yann.ponty@lix.polytechnique.fr}}

### 2013-01-09 Recomb/methods_RECOMB.tex
format llncs

\subsection{Probabilistic model}
Let $\Omega$ be an gap-free RNA alignment sequence, $S$ its associated secondary structure, 
then any sequence $s$ is assigned a probability proportional to its Boltzmann factor
\begin{align*}
  \mathcal{B}(s) &= e^\frac{-\PE{s}}{RT}, &&\text{with}&\PE{s}&:=\alpha\cdot\ES(s,S)+(1-\alpha)\cdot\EI(s,S,\Omega),
\end{align*}
where $R$ is the Boltzmann constant, $T$ the temperature in Kelvin, $\ES(s)$ and $\EI(s,S,\Omega)$ 
are the free-energy and isostericity contributions respectively (further described below), and $\alpha\in[0,1]$ is an arbitrary parameter that sets the relative weight for both contributions.

\subsubsection{Energy contribution}
The free-energy contribution in our pseudo-energy model corresponds to an additive stacking-pairs model, taking values from the Turner 2004 model retrieved from the NNDB~\cite{Turner2010}. Given a candidate sequence $s$ for a secondary structure $S$, the free-energy of $S$ on $s$ is given by
\begin{align*}
  \ES(s,S) = \sum_{\substack{(i,j)\to (i',j')\in S\\ \text{stacking pairs}}}\ES^{\beta}_{s_is_j\to s_{i'}s_{j'}} 
\end{align*}
where $\ES^{\beta}_{ab\to a'b'}$ is set to $0$ if $ab=\varnothing$ (no base-pair to stack onto), the tabulated free-energy of stacking pairs $(ab)/(a'b')$ in the Turner model if available, or $\beta\in[0,\infty]$ for non-Watson-Crick/Wobble entries (i.e. not in $\{\Gb\Ub,\Ub\Gb,\Cb\Gb,\Gb\Cb, \Ab\Ub\text{ or }\Ub\Ab\}$). This latter parameter allows to choose whether to simply penalize invalid base pairs, or forbid them altogether ($\beta = \infty$).
The loss of precision due to this simplification of the Turner model remains reasonable since the targeted secondary structure is fixed 
(e.g. multiloops do not account for base-specific contributions). Furthermore, it greatly eases the design and implementation of dynamic-programming equations. 
\subsubsection{Isostericity contribution}
The concept of isostericity score~\cite{Stombaugh2009} is based on the geometric discrepancy (superimposability) of two base-pairs, using individual additive contributions computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Let $s$ be a candidate sequence for a secondary structure $S$, given in the context of a gap-free RNA alignment $\Omega$,  we define the contribution of the isostericity to the pseudo-energy as
\begin{align*}
  \ES(s,S,\Omega) &= \sum_{\substack{(i,j)\in S\\ \text{pairs}}}\EI^{\Omega}_{(i,j),s_i s_j}, & \text{where}&& 	\EI^{\Omega}_{(i,j),ab}:=
	\frac{
		\sum_{s'\in\Omega}
			\text{\ISO}((s_i',s_j'),(a,b))}
%-		\left(			\text{\ISO}((s_i',s_j'),(s_i,s_j))				\right)	
{		
		|\Omega|
	}
\end{align*}
is the average isostericity of a base-pair in the candidate sequence, compared with the reference alignment.
The $\ISO$ function uses the {Watson-Crick/Watson-Crick} cis isostericity matrix computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Isostericity scores range between $0$ and $9.7$, with $0$ being assigned to a perfect isostericity, and a penalty of $10$ is used for missing entries.
The isostericity contribution will favor exponentially sequences that are likely to adopt a similar local conformation as the sequences contained in the alignment.

\subsection{Mutational profile of sequences}

@@

\subsubsection{Combining Inside and Outside values into point-wise mutations probabilities}
By construction, the partition function over all sequences at exactly $m$ mutations of $s$ can 
be described in function of the \emph{inside} term as $\Z{0,n-1}{m}{X,X}$,
 for any nucleotide $X\in B$ or
in function of the \emph{outside} term, for any unpaired position $k$:
$$
	\Z{0,n-1}{m}{X,X}
	\equiv
	\sum_{\substack{a\in \B,\\ \Kron_{a,s[k]}\le m}}	
	\Y{k-1,k+1}{m-\Kron_{a,s[k]}}{a,a}
$$

@@

\subsection{Complexity considerations}
Equations~\eqref{eq:Z_rec} and~\eqref{eq:Y_rec} can be computed using dynamic programming. Namely, the $\mathcal{Z}^{*}_{*}$ and $\mathcal{Y}^{*}_{*}$ terms are computed starting from smaller values of $m$ and interval lengths, memorizing the results as they become available to ensure constant-time access during later stages of the computation. Furthermore, energy terms $E(\cdot)$ can be accessed in constant time thanks to a simple precomputation (not described)  of the isostericity contributions in $\Theta(n\cdot|\Omega|)$. Computing any given term therefore requires $\Theta(m)$ operations.

>>>>

\subsection{Probabilistic Model}
Let $\Omega$ be an gap-free RNA alignment sequence, $S$ its associated secondary structure, 
then any sequence $s$ is assigned a probability proportional to its Boltzmann factor
\begin{align*}
  \mathcal{B}(s) &= e^\frac{-\PE{s}}{RT}, &&\text{with}&\PE{s}&:=\alpha\cdot\ES(s,S)+(1-\alpha)\cdot\EI(s,S,\Omega),
\end{align*}
where $R$ is the Boltzmann constant, $T$ the temperature in Kelvin, $\ES(s)$ and $\EI(s,S,\Omega)$ 
are the free-energy and isostericity contributions respectively (further described below), and $\alpha\in[0,1]$ is an arbitrary parameter that sets the relative weight for both contributions.

\subsubsection{Energy Contribution}
The free-energy contribution in our pseudo-energy model corresponds to an additive stacking-pairs model, taking values from the Turner 2004 model retrieved from the NNDB~\cite{Turner2010}. Given a candidate sequence $s$ for a secondary structure $S$, the free-energy of $S$ on $s$ is given by
\begin{align*}
  \ES(s,S) = \sum_{\substack{(i,j)\to (i',j')\in S\\ \text{stacking pairs}}}\ES^{\beta}_{s_is_j\to s_{i'}s_{j'}} 
\end{align*}
where $\ES^{\beta}_{ab\to a'b'}$ is set to $0$ if $ab=\varnothing$ (no base-pair to stack onto), the tabulated free-energy of stacking pairs $(ab)/(a'b')$ in the Turner model if available, or $\beta\in[0,\infty]$ for non-Watson-Crick/Wobble entries (i.e. not in $\{\Gb\Ub,\Ub\Gb,\Cb\Gb,\Gb\Cb, \Ab\Ub\text{ or }\Ub\Ab\}$). This latter parameter allows to choose whether to simply penalize invalid base pairs, or forbid them altogether ($\beta = \infty$).
The loss of precision due to this simplification of the Turner model remains reasonable since the targeted secondary structure is fixed 
(e.g. multiloops do not account for base-specific contributions). Furthermore, it greatly eases the design and implementation of dynamic-programming equations. 
\subsubsection{Isostericity Contribution}
The concept of isostericity score~\cite{Stombaugh2009} is based on the geometric discrepancy (superimposability) of two base-pairs, using individual additive contributions computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Let $s$ be a candidate sequence for a secondary structure $S$, given in the context of a gap-free RNA alignment $\Omega$,  we define the contribution of the isostericity to the pseudo-energy as
\begin{align*}
  \ES(s,S,\Omega) &= \sum_{\substack{(i,j)\in S\\ \text{pairs}}}\EI^{\Omega}_{(i,j),s_i s_j}, & \text{where}&& 	\EI^{\Omega}_{(i,j),ab}:=
	\frac{
		\sum_{s'\in\Omega}
			\text{\ISO}((s_i',s_j'),(a,b))}
%-		\left(			\text{\ISO}((s_i',s_j'),(s_i,s_j))				\right)	
{		
		|\Omega|
	}
\end{align*}
is the average isostericity of a base-pair in the candidate sequence, compared with the reference alignment.
The $\ISO$ function uses the {Watson-Crick/Watson-Crick} cis isostericity matrix computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Isostericity scores range between $0$ and $9.7$, with $0$ being assigned to a perfect isostericity, and a penalty of $10$ is used for missing entries.
The isostericity contribution will favor exponentially sequences that are likely to adopt a similar local conformation as the sequences contained in the alignment.

\subsection{Mutational Profile of Sequences}

@@

\subsubsection{Combining Inside and Outside Values into Point-Wise Mutations Probabilities}
By construction, the partition function over all sequences at exactly $m$ mutations of $s$ can 
be described in function of the \emph{inside} term as $\Z{0,n-1}{m}{X,X}$,
 for any nucleotide $X\in B$ or
in function of the \emph{outside} term, for any unpaired position $k$:
$$
	\Z{0,n-1}{m}{X,X}
	\equiv
	\sum_{\substack{a\in \B,\\ \Kron_{a,s[k]}\le m}}	
	\Y{k-1,k+1}{m-\Kron_{a,s[k]}}{a,a}
$$

@@

\subsection{Complexity Considerations}
Equations~\eqref{eq:Z_rec} and~\eqref{eq:Y_rec} can be computed using dynamic programming. Namely, the $\mathcal{Z}^{*}_{*}$ and $\mathcal{Y}^{*}_{*}$ terms are computed starting from smaller values of $m$ and interval lengths, memorizing the results as they become available to ensure constant-time access during later stages of the computation. Furthermore, energy terms $E(\cdot)$ can be accessed in constant time thanks to a simple precomputation (not described)  of the isostericity contributions in $\Theta(n\cdot|\Omega|)$. Computing any given term therefore requires $\Theta(m)$ operations.

### 2013-01-09 Recomb/results_RECOMB.tex
Revamped part of method section One more hunt for typos
format llncs

\subsection{Implementation}
The software was implemented in Python2.7 using the \textit{mpmath}~\cite{mpmath} library
for  arbitrary floating point precision. The source code is freely available at \verb+https://github.com/McGill-CSB/RNApyro+.

The time benchmarks were done on a MacMini 2010, 2.3GHz dual-core Intel Core i5, 8GB of RAM.
Since applications of \RNApyro implies a need for
 efficiency and scalability, we present in Table~\ref{tab:time} typical runtimes required to compute the probabilities for  every nucleotide at every positions for a vast set of parameters. For those test
 both the sequences and the target secondary structure were generated at random.

\begin{table}[b]
\begin{center}
\begin{tabular}{lccc}
$|s|$&\multicolumn{3}{c}{Number of mutations}\\\cline{2-4}
		 			  & 6   &  12  & 24\\\cline{2-4}
100  				& 35s  & 238s & 1023s\\
300  			& 135s & 594s &2460s\\\cline{2-4}
		 						& 25   &    &	50		\\\cline{2-4}
500         & 5400s&       &  21003s    \\
\end{tabular}
\end{center}
\caption{Time to compute all probabilities. The first column indicates the length and  the column indexes indicate the number
 of mutations. $\alpha$ is
set at $0.5$,  $\beta$ to $1.5$ and $|\Omega|=44$.}
\label{tab:time}
\end{table}

\subsection{Error correction in 5s rRNA}

@@

\begin{figure}
\begin{subfigure}[b]{0.3\textwidth}
\includegraphics[width=1.2\textwidth]{figures/ROC_6.pdf}
\caption{6 mutations}
\label{fig:ROC6mut}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.3\textwidth}
\includegraphics[width=1.2\textwidth]{figures/ROC_12.pdf}
\caption{12 mutations}
\label{fig:ROC12mut}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.3\textwidth}
\includegraphics[width=1.2\textwidth]{figures/ROC_24.pdf}
\caption{24 mutations}
\label{fig:ROC24mut}
\end{subfigure}
\caption{Performance of error-correction. Subfigures accuracy with under-estimated error rate (6 mutations), exact estimates (12 mutations) and over estimates 
(24 mutations). We also analyze the impact of the parameter $\alpha$ distributing the weights of stacking pair energies vs isostericity scores and use values 
ranging of $\alpha=\{0,0.5,0.8,1.0\}$. The AUC is indicated in the legend of the figures. Each individual ROC curve represent the average performance over the 10 experiments.}
\label{fig:ROCall}
\end{figure}

Our data demonstrate that our algorithm exhibits interesting performance for error-correction applications. First, the AUC values (up to $0.86$) indicate that a
signal has been successfully extracted. This result has been achieved with errors in loop regions (i.e. without base pairing information) and thus suggests
that correction rates in structured regions (i.e. base paired regions) could be even higher. Next, the optimal values of $\alpha$ tend to be close to $0.0$. This 
finding suggests that at this point the information issued from the stacking energy is currently modest. However, specific examples showed improved performance
using this energy term. Further studies must be conducted to understand how to make the best use it. Finally, our algorithm seems robust to the number of
mutations considered. Indeed, good AUC values are achieved with low estimates of the number of errors in the sequence (c.f. 50\% of the errors  in
Fig.~\ref{fig:ROC6mut}) and with large  values (c.f.~200\% of the errors  in Fig.~\ref{fig:ROC24mut}). It is worth noting that scoring scheme with larger weight on
the isostericity scores (low $\alpha$ values) seem more robust to under- and over-estimate of the number of errors.

>>>>

\subsection{Implementation}
The software was implemented in Python2.7 using the \textit{mpmath}~\cite{mpmath} library
for  arbitrary floating point precision. The source code is freely available at: \\\hfill\verb+https://github.com/McGill-CSB/RNApyro+.

The time benchmarks were done on a MacMini 2010, 2.3GHz dual-core Intel Core i5, 8GB of RAM.
Since applications of \RNApyro implies a need for
 efficiency and scalability, we present in Table~\ref{tab:time} typical runtimes required to compute the probabilities for  every nucleotide at every positions for a vast set of parameters. For those tests,
 both the sequences and the target secondary structure were generated at random.

\begin{table}[t]
\begin{center}
\begin{tabular}{lccc}
$|s|$&\multicolumn{3}{c}{Number of mutations}\\\cline{2-4}
		 			  & 6   &  12  & 24\\\cline{2-4}
100  				& 35s  & 238s & 1023s\\
300  			& 135s & 594s &2460s\\\cline{2-4}
		 						& 25   &    &	50		\\\cline{2-4}
500         & 5400s&       &  21003s    \\
\end{tabular}
\end{center}
\caption{Time to compute all probabilities. The first column indicates the length and  the column indexes indicate the number
 of mutations. $\alpha$ is
set at $0.5$,  $\beta$ to $1.5$ and $|\Omega|=44$.}
\label{tab:time}
\end{table}

\subsection{Error Correction in 5s rRNA}

@@

\begin{figure}
\begin{center}
	\includegraphics[width=\textwidth]{subfigs_perform.png}
\end{center}
%\begin{subfigure}[b]{0.3\textwidth}
%\includegraphics[width=1.2\textwidth]{figures/ROC_6.pdf}
%\caption{6 mutations}
%\label{fig:ROC6mut}
%\end{subfigure}
%\hfill
%\begin{subfigure}[b]{0.3\textwidth}
%\includegraphics[width=1.2\textwidth]{figures/ROC_12.pdf}
%\caption{12 mutations}
%\label{fig:ROC12mut}
%\end{subfigure}
%\hfill
%\begin{subfigure}[b]{0.3\textwidth}
%\includegraphics[width=1.2\textwidth]{figures/ROC_24.pdf}
%\caption{24 mutations}
%\label{fig:ROC24mut}
%\end{subfigure}
\caption{Performance of error-correction. Subfigures accuracy with under-estimated error rate (6 mutations), exact estimates (12 mutations) and over estimates 
(24 mutations). We also analyze the impact of the parameter $\alpha$ distributing the weights of stacking pair energies vs isostericity scores and use values 
ranging of $\alpha=\{0,0.5,0.8,1.0\}$. The AUC is indicated in the legend of the figures. Each individual ROC curve represent the average performance over the 10 experiments.}
\label{fig:ROCall}
\end{figure}

Our data demonstrate that our algorithm exhibits interesting performance for error-correction applications. First, the AUC values (up to $0.86$) indicate that a
signal has been successfully extracted. This result has been achieved with errors in loop regions (i.e. without base pairing information) and thus suggests
that correction rates in structured regions (i.e. base paired regions) could be even higher. Next, the optimal values of $\alpha$ tend to be close to $0.0$. This 
finding suggests that at this point the information issued from the stacking energy is currently modest. However, specific examples showed improved performance
using this energy term. Further studies must be conducted to understand how to make the best use it. Finally, our algorithm seems robust to the number of
mutations considered. Indeed, good AUC values are achieved with low estimates of the number of errors in the sequence (c.f. 50\% of the errors  in
Fig.~\ref{fig:ROCall}(a)) and with large  values (c.f.~200\% of the errors  in Fig.~\ref{fig:ROCall}(c)). It is worth noting that scoring scheme with larger weight on
the isostericity scores (low $\alpha$ values) seem more robust to under- and over-estimate of the number of errors.
